---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

```{r}
library(readxl)
library(here)
library(tidyverse)
library(ggbeeswarm)

mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))

tcsa <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
  #omit very large TCSA, could be measurement error
  filter(min_tcsa <= 250) %>%
  # group rare raw materials
  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
  mutate(site_name_top = fct_lump_min(Site_name, 5))
```


```{r}
# all artefacts overall
ggplot (tcsa,
        aes(1, min_tcsa)) +
  # show regions for tools from Table 3 of Lomabrd 2021
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
    annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 4,
           ymax = 18,
           fill = "grey80") +
      annotate("text", 
               label = "Poisoned arrowheads",
           x = 1.3,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 18,
           ymax = 53,
           fill = "grey50") +
      annotate("text", 
               label = "Arrowheads",
           x = 1.3,
           y = (53-13) /2 + 13) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 53,
           ymax = 76,
           fill = "grey90") +
      annotate("text", 
               label = "Dart tips",
           x = 1.3,
           y = (76-53) /2 + 53 ) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 76,
           ymax = 98,
           fill = "grey70") +
      annotate("text", 
               label = "Javelin",
           x = 1.3,
           y = (98-76) /2 + 76 ) +
     annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 98,
           ymax = 230,
           fill = "grey95") +
      annotate("text", 
               label = "Thrusting spear",
           x = 1.3,
           y = (230-98) /2 + 98 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 3) +

  theme_bw()
```




```{r}
#### histogram for min tcsa to prepare for exploring the data
all <- ggplot(tcsa,
              aes(min_tcsa)) +
  geom_histogram() +
  theme_bw()
```


## Raw materials

```{r}
# Facet by raw materials
all +  facet_wrap(~raw_materials_top,
                  scales = "free_y",
                  ncol = 1)
```

```{r}
# remove NAs
tcsa_mass_no_na <-
  tcsa %>%
 drop_na(`Weight(g)`)

# how does the size-tcsa relationship vary by raw material?
ggplot(tcsa_mass_no_na,
       aes(`Weight(g)`,
           min_tcsa,
           colour = raw_materials_top)) +
  geom_point() +
  theme_bw()
```

## Variation by site

```{r}
# Facet by site
all +  facet_wrap(~site_name_top,
                 scales = "free_y",
                  ncol = 1)
```


## Variation by artefact size

```{r}
# Facet by size class

# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
library(Ckmeans.1d.dp)

# find clusters
result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)

tcsa_mass_no_na <-
tcsa_mass_no_na %>%
  mutate(mass_cluster = factor(result$cluster))

ggplot (tcsa_mass_no_na,
        aes(1, `Weight(g)`)) +
  geom_boxplot() +
  geom_quasirandom(aes(colour = mass_cluster),
                   alpha = 0.4,
                   size = 2) +
  theme_bw()

ggplot(tcsa_mass_no_na,
       aes(min_tcsa)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~mass_cluster,
             ncol = 1,
             scales = "free_y")
```


# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
