---
title: "Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea"
author:
  - Gayoung Park: 
      email: gayoungp@uw.edu
      institute: [UofW]
      correspondence: true
  - Marlize Lombard:
      email: mlombard@uj.ac.za
      institute: [UofJ]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofW]
      correspondence: false
  - Donghee Chong:
      email: okidok2@khu.ac.kr
      institute: [UofK]
      correspondence: false
      
institute:
  - UofW: Department of Anthropology, University of Washington, Seattle. USA
  - UofJ: Palaeo-Research Institute, University of Johannesburg, South Africa
  - UofK: Kyung Hee University, South Korea

date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: false
      global_numbering: true
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

<!-- hide these heading for now:

# Introduction

# Background

# Methods

# Results

-->

```{r}
library(readxl)
library(here)
library(tidyverse)
library(ggbeeswarm)
library(ggplot2)

mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))

tcsa <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
  #omit very large TCSA, could be measurement error
  filter(min_tcsa <= 250) %>%
  # group rare raw materials
  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
  mutate(site_name_top = fct_lump_min(Site_name, 5))

```

```{r, map, fig.cap="Korean Paleolithic sites mentioned in this study."}
# Facet by raw materials
korean_archaeological_site_locations <-
  read_excel(here::here("analysis/data/raw_data/TCSA_site_locations.xlsx"))

korean_archaeological_site_locations <-
  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")

library(ggmap)
library(ggrepel)
library(legendMap)

map <-
  get_stamenmap(bbox = c(left = 124,
                         bottom = 33,
                         right = 	130,
                         top = 39),
                zoom = 10)

# map with site names next to the points
map_of_sites <-
  ggmap(map)  +
  geom_point(data = korean_archaeological_site_locations,
             aes(long_dd ,
                 lat_dd),
             colour = "red",
             size = 2) +
  # label points with site name
  geom_text_repel(data = korean_archaeological_site_locations,
                  aes(long_dd ,
                      lat_dd,
                      label = Site_name),
                  size = 2,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_minimal(base_size = 6) +
  labs(x = "Longitute",
       y = "Latitude") +
  legendMap::scale_bar(
    # edit these numbers to select a suitable location
    # for the scale bar where it does not cover
    # important details on the map
    lon = 124.5,
    lat = 33.5,
    legend_size = 2,
    # distance of one section of scale bar, in km
    distance_lon = 50,
    # height of the scale bar, in km
    distance_lat = 5,
    # distance between scale bar and units, in km
    distance_legend = 20,
    # units of scale bar
    dist_unit = "km",
    # add the north arrow
    orientation = TRUE,
    # length of N arrow, in km
    arrow_length = 60,
    # distance between scale bar & base of N arrow, in km
    arrow_distance = 50,
    # size of letter 'N' on N arrow, in km
    arrow_north_size = 10) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())


# inset world map, from
# https://stackoverflow.com/a/62104396/1036500
library(maps)
library(magrittr)
library(maptools)
library(raster)
library(ggplot2)

#Defining a general CRS
mycrs <- "+proj=longlat +datum=WGS84 +no_defs"

#Using the original maps package, then converting map into SpatialPolygons object
world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))

#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
while(rgeos::gIsValid(world)==FALSE){
  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
}

#Dissolving polygon's limits
world <- raster::aggregate(world)

worldplot <-
  ggplot() +
  geom_polygon(data = world,
               aes(x=long, y=lat,
                   group=group),
               fill='NA',
               color='black',
               size=0.1) +
  annotate("rect",
           xmin = 118,
           xmax = 137,
           ymin = 28,
           ymax = 47,
           fill = NA,
           colour = "red",
           size = 0.5
  ) +
  theme_void() +
  coord_fixed(1) +
  labs(x = "", y = "") +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

# combine site map and inset map
library(cowplot)
ggdraw() +
  draw_plot(map_of_sites) +
  draw_plot(plot = worldplot,
            x = 0.6, # x location of inset placement
            y = -0.09, # y location of inset placement
            width = .4, # Inset width
            height = .35, # Inset height
            scale = 0.5 # Inset scale
  ) +
  theme(panel.background = element_rect(fill='white', colour="white"),
        plot.background = element_rect(fill='white', colour="white"))

```

```{r, all-stemmed-points, fig.width= 4.5, fig.height=4, fig.cap="Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types from Lombard (2021)."}
# all artefacts overall

point_function_group_label_y <- 0.75
point_function_group_label_size <- 2

ggplot(tcsa,
        aes(0.3, 
            min_tcsa)) +
  # show regions for tools from Table 3 of Lomabrd 2021
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned\narrowheads",
               hjust = 0,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 13,
           ymax = 53,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Arrowheads",
               hjust = 0,
           x = point_function_group_label_y,
           y = (53-13) /2 + 13) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 ) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 44,
           ymax = 98,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin",
               hjust = 0,
           x = point_function_group_label_y,
           y = (98-44) /2 + 44 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 88,
           ymax = 230,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Thrusting\nspear",
               size = point_function_group_label_size,
               hjust = 0,
           x = point_function_group_label_y,
           y = (230-88) /2 + 88 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 2.3) +
  coord_cartesian(xlim = c(-0.1,1)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")
```

Notes on Figure \@ref(fig:all-stemmed-points):

1) The mean TCSA of the Korean stemmed point is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`. Overall we see wide variation of TCSA values in this boxplot.   
2) According to Lombard's (2021) TCSA ranges, Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
3) Poison arrowheads seem to be absent from these Korean assemblages


```{r}

### create column for "n = xx "
tcsa_raw <- tcsa %>% 
  group_by(raw_materials_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
  dplyr::select(min_tcsa,
         raw_materials_top, 
       label_raw,
       n)   %>% 
  distinct_all()

tcsa_site <- tcsa %>% 
  group_by(site_name_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
  select(min_tcsa,
         site_name_top, 
         label_site,
         n) %>% 
  distinct_all()
```

## Variation by raw material

```{r, raw-materials, fig.width= 4.5, fig.height=4, fig.cap="Histograms of TCSA values by lithic raw material."}
# Facet by raw materials
 ggplot(tcsa_raw,
              aes(min_tcsa, 
                  fill = n)) +
  geom_histogram() +
  theme_bw(base_size = 6) +
  labs(x = "TCSA", 
       y = "Count of artefacts") +
  facet_wrap(~label_raw,
                  scales = "free_y",
                  ncol = 1)  +
  scale_fill_viridis_c(name = "Number\nof artefacts")
 
```

Notes on Figure \@ref(fig:raw-materials):

1) About half of all the stemmed points made out of shale. 
2) Acidic volcanic rocks tend to have skewed lower TCSA values. 
3) But overall, there is no clear pattern of different TCSA values by raw material.


## Variation by site

```{r, sites, fig.width= 4.5, fig.height=4, fig.cap="Histograms of TCSA values by archaeological site."}
# Facet by site
 ggplot(tcsa_site,
              aes(min_tcsa, 
                  fill = n)) +
  geom_histogram() +
  theme_bw(base_size = 6) +
  labs(x = "TCSA", 
       y = "Count of artefacts") +
  facet_wrap(~label_site,
                 scales = "free_y",
                  ncol = 1) +
    scale_fill_viridis_c(name = "Number\nof artefacts")
```

Notes on Figure \@ref(fig:sites):

1) We computed TCSA for sites that include >= 5 stemmed points.
2) Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points for a variety of different purposes. 
3) Nosanri and Sibuk have similarly small TCSA values, perhaps these sites had similar functions? Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but different distribution of TCSA values, so perhaps Wolpyeng and Yongsujaeul had a different role in the foragers' land use system compared to Nosanri and Sibuk.

## Variation by artefact size

We wanted to check the relationship between TCSA and size. We use weight to represent overall tool size. 

```{r, size-tsca, fig.width= 4.5, fig.height=4, fig.cap="Artefact size and TSCA values by lithic raw material"}
# how does the size-tcsa relationship vary by raw material?

# remove NAs
tcsa_mass_no_na <-
  tcsa %>%
 drop_na(`Weight(g)`)

ggplot(tcsa_mass_no_na,
       aes(`Weight(g)`,
           min_tcsa,
           colour = raw_materials_top)) +
  geom_point() +
    labs(x = "Weight", 
       y = "TCSA",
       colour='Raw material') +
  geom_smooth(method=lm, se=FALSE, size = 0.4)+
  theme_bw(base_size = 6) +
  theme(
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 7),
  legend.position = c(.8,.3)
  )
```

Notes on Figure \@ref(fig:size-tsca):

1) TCSA and artefact weight have a positive relationship.
2) The relationship between TCSA and artefact weight varies by raw material, perhaps due to variation in the density of different lithic raw materials
3) Shale stemmed points tend to have TCSA values that are strongly correlated with artefact size, while acidic volcanic points have TCSA values that vary over a very limited size range, suggesting size is less influential. 

```{r, size-class, fig.width= 4.5, fig.height=4, fig.cap=""}
# Facet by size class

# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
library(Ckmeans.1d.dp)

# find clusters
result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)

tcsa_mass_no_na <-
tcsa_mass_no_na %>%
  mutate(mass_cluster = factor(result$cluster))

plot_mass_cluster <- 
ggplot (tcsa_mass_no_na,
        aes(1, 
            `Weight(g)`)) +
  geom_boxplot(outlier.shape = NA) +
  geom_quasirandom(aes(shape = mass_cluster),
                   alpha = 0.4,
                   size = 2.5) +
  labs(shape='Cluster',
       y = "Weight (g)") +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        )
```


```{r, artefact-size-cluster,fig.width= 4.5, fig.height=4, fig.cap="A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."}
plot_tcsa_mass_cluster <- 
ggplot (tcsa_mass_no_na,
        aes(0.3, min_tcsa)) +
  geom_boxplot(outlier.shape = NA) +
  geom_quasirandom(aes(shape = mass_cluster),
                   alpha = 0.4,
                   size = 2.5) +
  labs(shape='Cluster') +
  theme_bw(base_size = 6) +
  # show regions for tools from Table 3 of Lomabrd 2021
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
    annotate("rect", 
           xmin = 0.1,
           xmax = 0.2,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned\narrowheads",
               hjust = 0,
           x = point_function_group_label_y,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 0.2,
           xmax = 0.3,
           ymin = 13,
           ymax = 53,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Arrowheads",
               hjust = 0,
           x = point_function_group_label_y,
           y = (53-13) /2 + 13) +
      annotate("rect", 
           xmin = 0.3,
           xmax = 0.4,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = point_function_group_label_y,
           y = (76-40) /2 + 40 ) +
      annotate("rect", 
           xmin = 0.4,
           xmax = 0.5,
           ymin = 44,
           ymax = 98,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin",
               hjust = 0,
           x = point_function_group_label_y,
           y = (98-44) /2 + 44 ) +
     annotate("rect", 
           xmin = 0.5,
           xmax = 0.6,
           ymin = 88,
           ymax = 230,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Thrusting\nspear",
               size = point_function_group_label_size,
               hjust = 0,
           x = point_function_group_label_y,
           y = (230-88) /2 + 88 ) +
  coord_cartesian(xlim = c(-0.1,1.2)) +
  theme_bw(base_size = 6) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
        ) +
  labs(y = "TCSA")

library(cowplot)
plot_grid(plot_mass_cluster + guides(shape = "none"),
          plot_tcsa_mass_cluster,
          labels = "AUTO",
          rel_widths = c(0.5, 1))
```

Notes on Figure \@ref(fig:artefact-size-cluster):

1) Univariate cluster analysis of artefacts by weight reveals three groups or clusters 
2) Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.  
3) We can assumed that artifacts in different clusters might be made for different purposes.


<!-- hide these for now

# Discussion

# Conclusion

# Acknowledgements

--> 


<!-- The following line inserts a page break  -->



# CRediT authorship contribution statement
Gayoung Park: Software, Validation, Formal analysis, Resources, Data curation,  Writing - original draft, Writing - Review & Editing, Visualization, Project administration.

Marlize Lombard: Conceptualization, Methodology, Writing - Review & Editing, Supervision

Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision

Donghee Chong: Data curation, Review & Editing

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
